Last glacial cycle glacier erosion potential in the Alps

### 2021-01-28 articles/erosion/alps-erosion.tex
Rework/draft discussion on age of the landscape.

    Deglaciation periods, characterised by a warmer climate and above-balance
    glacier extent, yield the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). This
    result is compatible with the field-based studies, which have evidenced
    increased glacier erosion and pro-glacial sedimentation during deglaciation
    periods \ref, including the ongoing glacier retreat \aref. Yet this
    increase has often been attributed to higher meltwater availability,
    promoting river erosion, enhanced glacier sliding, and hydrofracturing
    (plucking) \aref.

@@

    The results also indicate that during stages of extensive glaciation, much
    of the intra-Alpine region experiences insignificant erosion rates. For
    instance, from 35 to 18\,ka BP, terrain with elevations above 1000\,m are
    modelled to experiences virtually no glacier erosion. This would imply that
    the spectacular glacial valleys found throughout the northwestern Alps do
    not date from the LGM or similar periods, but from periods of intermediate,
    and likely shrinking, ice cover.

>>>>

    Deglaciation periods, characterised by a warmer climate and above-balance
    glacier extent, yield the highest domain-integrated erosion rates of the
    glacial cycle (Figs.~\ref{fig:cumulative} and~\ref{fig:evolution}). This
    result is compatible with the field-based studies, which have evidenced
    increased glacier erosion and pro-glacial sedimentation during deglaciation
    periods \aref, including the ongoing glacier retreat \aref. Yet this
    increase has often been attributed to higher meltwater availability,
    promoting river erosion, enhanced glacier sliding, and hydrofracturing
    (plucking) \aref.

@@

    In many places throughout the Alps, significant modelled cumulative erosion
    occurs far inside the Alpine valleys (Fig.~\ref{fig:cumulative}). Yet
    instantaneous erosion rates show a time-transgressive radial pattern
    (Fig.~\ref{fig:transects}), so that high-altitude rapid erosion is not
    modelled to date from expansive glaciation stages, but instead to stages of
    intermediate glacier cover (Fig.~\ref{fig:hypsogram}).

    For instance, low-elevation piedmont over-deepened troughs such as Lake
    Constance (Fig.~\ref{fig:landscape}), Lake Geneva and Lake Maggiore, are
    here modelled to be the product of extensive, Last Glacial Maximum-like
    glaciations. Alpine terrain with elevations above 1000\,m are modelled to
    experience virtually no erosion from 35 to 18\,ka BP
    (Fig.~\ref{fig:hypsogram}), so that many of the most spectacular,
    intermediate-elevation glacial valley of the Alps such as Lauterbrunnental
    (Fig.~\ref{fig:hypsogram}), Haslital (Upper Aare) or Bout du Monde (Upper
    Giffre). While the validity of the model results at high elevation are
    limited by the 1-km resolution and the shallow-ice physics
    \citep{Imhof.etal.2019}, the lack of high-elevation rapid erosion for much
    of the modelled glacial cycle implies that the numerous Alpine mountain
    cirques such as Ch\"uebodensee (Fig.~\ref{fig:landscape}) would date from
    temperate periods with limited ice-cover such as the present day.

    Finally, the time-transgressive nature of the modelled glacial erosion
    rates hint at a new explanation for the Alpine glacier trim-line and lack
    of cold-based glaciation evidence. Due to their geographic location in the
    mid-latitudes, and their steep topographic gradient, the Alps appear as a
    mountain range that hosted glaciers of various sizes throughout the varying
    climate of the Quaternary, resulting in a transgressive localization of
    glacier erosion throughout their elevational range, from the piedmont
    during glacial maxima, to the highest valleys during interglacial periods.
    %The Alpine trim-line, in this case, would neither correspond to the upper
    %reach of the Last Glacial Maximum glaciers \aref, nor to an englacial
    %temperate-ice boundary \aref, but to an upper limit of erosion from
    %retreating glaciers on steep terrain.
